---
title: "732A90 - Exam - Task 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1

## Inverse CDF

1. Calculate the normalization constant:

\begin{align}
P(x) 
&= \int_0^1 a^x dx\\
&= \int_0^1 e^{xlog(a)} dx\\
&= \left[ \frac{1}{log(a)} e^{x log(a)}\right]_0^1 \\
&= \frac{a - 1}{log(a)} \\
&=> c = \frac{log(a)}{a - 1}
\end{align}

2. Calculate the CDF

\begin{align}
P(y) 
&= \int_0^y ca^x dx\\
&= c\int_0^y e^{xlog(a)} dx\\
&= c\left[ \frac{1}{log(a)} e^{x log(a)}\right]_0^y \\
&= \frac{c}{log(a)} (e^{ylog(a)} - 1)
\end{align}

3. Inverse it

\begin{align}
y = \frac{1}{a-1} (e^{xlog(a)} - 1) \\
x = \frac{1}{a-1} (e^{ylog(a)} - 1)\\
\frac{log(x(a-1)+1)}{log(a)} = y
\end{align}

```{r}
inverse_cdf <- function(x, a) {
  log(x * (a - 1) + 1) / log(a)
}

u <- runif(1000)
res_11 <- sapply(u, inverse_cdf, a = 0.1)

a <- 0.1
c <- log(a) / (a-1)
x <- seq(0.01, 0.99, 0.01)
y <- a^x * c

hist(res_11, freq = FALSE, breaks = 40,
     xlim = c(0,1 ), ylim = c(0, 3),
     xlab = "x", ylab = "Density",
     main = "Histrogram inverse CDF sampling",
     col = "grey")
lines(density(res_11), col = "blue", lwd = 2)
lines(x, y, col = "red", lwd = 2)
legend(x = 0.75, y = 3,
       legend = c("Sampled values", "Sample Density", "Real Values"),
       col = c("grey", "blue", "red"),
       pch = c(16, 16, 16),
       cex = 0.8)
```

Sampled values look like the values from $p(y)$. The blue line will converge towards the red line if the sample size increases.